---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fort ~ fast orthogonal random transforms in R

<!-- badges: start -->

<!-- badges: end -->

The `fort` package provides convenient access to fast structured random linear transforms that are (at least approximately) orthogonal or semi-orthogonal, and generally faster than matrix multiplication (in the style of the Fastfood Transform), implemented in C++ (via Rcpp).

Useful for algorithms that require or benefit from uncorrelated random projections, such as fast dimensionality reduction (e.g., Johnson-Lindenstrauss transform) or kernel approximation (e.g., random kitchen sinks).

## Installation

You can install the development version of `fort` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tomessilva/fort")
```

## Example

This is a basic example which shows you how to solve use `fort` in practice:

```{r example}
library(fort)

(fast_transform <- fort(4)) # fast orthogonal transform from R^4 to R^4
matrix_to_transform <- diag(4) # 4 x 4 identity matrix
(new_matrix <- fast_transform %*% matrix_to_transform) # transformed matrix
inverse_transform <- solve(fast_transform) # get inverse transform
(inverse_transform %*% new_matrix) # should approximately recover the identity matrix
```

