---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fort ~ fast orthogonal random transforms in R

<!-- badges: start -->

![](https://img.shields.io/badge/package-fort-blue?logo=r)

<!-- badges: end -->

The `fort` package provides convenient access to fast structured random linear transforms that are (at least approximately) orthogonal or semi-orthogonal, and generally faster than matrix multiplication (in the style of the Fastfood Transform), implemented in C++ (via Rcpp).

Useful for algorithms that require or benefit from uncorrelated random projections, such as fast dimensionality reduction (e.g., Johnson-Lindenstrauss transform) or kernel approximation (e.g., random kitchen sinks).

## Installation

You can install the development version of `fort` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tomessilva/fort")
```

## Example

This is a basic example which shows how to use `fort` in practice:

```{r example}
library(fort)

(fast_transform <- fort(4)) # fast orthogonal transform from R^4 to R^4

matrix_to_transform <- diag(4) # 4 x 4 identity matrix
(new_matrix <- fast_transform %*% matrix_to_transform) # transformed matrix

(inverse_transform <- solve(fast_transform)) # get inverse transform

round(inverse_transform %*% new_matrix, 12) # should recover the identity matrix
```

Here is a comparison of using a `fort` transform against a simple matrix multiplication, in terms of speed:

``` r 
library(fort)

matrix_to_transform <- diag(1024) # 1024 x 1024 identity matrix

fast_transform <- fort(1024) # fast orthogonal transform from R^1024 to R^1024
slow_transform <- as.matrix(fast_transform) # the same, but in matrix form

# time it takes for the fast transform
system.time(for (i in 1:100) test <- fast_transform %*% matrix_to_transform, gcFirst = TRUE)
#>   user  system elapsed 
#>   5.50    2.12    8.00

# time it takes for the equivalent slow transform (via matrix multiplication)
system.time(for (i in 1:100) test <- slow_transform %*% matrix_to_transform, gcFirst = TRUE)
#>   user  system elapsed 
#>  70.57    0.61   77.95 
```
